ASHA 2023 Listening Effort Data Analysis

### 2023-10-30 ASHA 2023 Analysis/data_analysis.Rmd
Streamlined code in rmd file

peak_pupil_df %>%
  ggplot() +
  aes(y = speaker,
      x = peak_pupil,
      color = speaker,
      fill = speaker) +
  geom_density_ridges(scale = 0.7, alpha = 0.5) +
  theme_bw() +
  scale_fill_manual(values = my_pal) +
  scale_color_manual(values = my_pal)

@@

As indicated from the model, adding trial order did not significantly improve the model fit to the data (chi-sq(1)m p = 0.52). This indicates that trial order is not an important predictor for perceived listening effort ratings and that ratings for the speakers were fairly consistent across trials.

@@

Based on the model output, adding the interaction term did not significantly improve model fit (chi-sq(1) = 0.70). The fixed effect estimate for the interaction term is also not significant. This indicates that effort ratings across the trials for the ALS speaker were not significantly different from each other.

@@

summary(per_effort_model_final)

@@

rm(m0, m1, m2, m3, m4, per_effort_des, perceived_effort_df, per_effort_model_final, model_0, model0)

@@

@@

rm(m0, m1, m2, m3, m4, pupil_plot, pupil_roi, data, control_bobyqa, my_pal)

@@

@@

# Ran into convergence issues, so added the nlminb optimizer to the model.
# Model had simgular fit issues, so removed the (trial_c|code) random effect.

m1 <- lmer(peak_pupil ~ 1 + (trial_c|subject), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

@@

m2 <- lmer(peak_pupil ~ trial_c + (trial_c|subject), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

@@

m3 <- lmer(peak_pupil ~ trial_c + speaker + (trial_c|subject) , 
              data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

@@

m4 <- lmer(peak_pupil ~ trial_c + speaker + trial_c*speaker + (trial_c|subject) , 
              data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

@@

@@

>>>>

perceived_effort_df %>%
  ggplot() +
  aes(y = speaker,
      x = effort_rating,
      color = speaker,
      fill = speaker) +
  geom_density_ridges(scale = 0.7, alpha = 0.5) +
  labs(x = "Perceived Listening Effort Rating", y = "Speaker") +
  theme_bw() +
   theme(legend.position = "none", aspect.ratio = 1) +
  scale_fill_manual(values = my_pal) +
  scale_color_manual(values = my_pal)

ggsave("Plots/Perceived_effort_dis.png", plot = last_plot(), height = 5, width = 5, units = "in")

@@

As indicated from the model, adding trial order did not significantly improve the model fit to the data (chi-sq(1) = 0.412, p = 0.52). This indicates that trial order is not an important predictor for perceived listening effort ratings and that ratings for the speakers were fairly consistent across trials.

@@

Based on the model output, adding the interaction term did not significantly improve model fit (chi-sq(1) = 0.14, p= 0.70). The fixed effect estimate for the interaction term is also not significant. This indicates that effort ratings across the trials for the ALS speaker were not significantly different from each other.

@@

@@

rm(m0, m1, m2, m3, m4, per_effort_des, perceived_effort_df, per_effort_model_final)

@@

sjPlot::tab_model(m4, file = "Final Model Outputs/final_pupil_dilation_model.html")

@@

rm(m0, m1, m2, m3, m4, pupil_plot, control_bobyqa)

@@

We first fit a fully unconditional model to the data to calculate ICC. The adjusted and unadjusted ICC is 0.364.

# Note: I ran into issues using glmmTMB, so used lme4 instead

@@

We added trial as random slopes for both the subject and code random effects. The model initially failed to converge, so an optimizer was added to the model. Then a singular fit warning was displayed, so the random slope of trial_c was removed from the code random effect.

# Model failed to converge, so added an optimizer
# Singular fit so removed random slope for code.

m1 <- lmer(peak_pupil ~ 1 + (trial_c|subject) + (1|code), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

Model 1 provided a better fit to the data (chi-sq(2) = 43.44, p < 0.001).

@@

Next, in order to control for trial order, trial was added to the model as a fixed effect.

m2 <- lmer(peak_pupil ~ trial_c + (trial_c|subject) + (1|code), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

Adding trial to the model as a fixed effect significantly improved model fit (chi-sq(1) = p<0.001). Based on the fixed effect estimate for trial order, peak pupil dilation decreased approximately -1.23 units on average with each trial.

@@

Next, we added the fixed effect of speaker to the model to test whether peak pupil dilation was, on average, higher for when listening to the ALS speaker compared to the control speaker.

m3 <- lmer(peak_pupil ~ trial_c + speaker + (trial_c|subject) + (1|code), 
           data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

Adding speaker to the model improved the model fit to the data (ch-sq(1), p <0.001). Controlling for trial order, listeners' peak pupil dilation, on average, was about 85.22 units higher compared to when they were listening to the control speaker. This indicates listeners were exerting more effort while listening to the ALS speaker compared to the control speaker.

@@

Finally, the interaction between trial and speaker was added to the model to test whether peak pupil dilation for the ALS speaker change at a different rate across trials.

# Ran into convergence problem, so removed the trial_c random slope for code

m4 <- lmer(peak_pupil ~ trial_c + speaker + speaker*trial_c + (trial_c|subject) + (1|code), 
              data = peak_pupil_df, control = lmerControl(
                           optimizer ='optimx', optCtrl=list(method='nlminb')))

@@

Based on the model output, adding the speaker*trial interaction did not significantly improve model fit to the data (chi-sq(1) = 0.06, p = .80). This indicates that when listening to the speaker with ALS, the listeners' peak pupil dilation declined at a similar rate as when they were listening to the control speaker.

@@

Based on the model comparisons, model 3 provided the best fit to the data. Therefore, the final model output for peak pupil dilation is provided below.

@@

rm(data, m0, m1, m2, m3, m4, peak_pupil_des, peak_pupil_df, pupil_roi, my_pal)
